---
title: "Preprocessing script for Hayashi 2018"
author: "Aditya Pratapa"
date: "`r Sys.Date()`"
output: 
  BiocStyle::html_document:
    toc: true
vignette: >
---

### MAKE SURE the workspace is scripts/

```{r}
script_path <- dirname(rstudioapi::getActiveDocumentContext()$path)
setwd(script_path)
```

### Load necessary libraries
```{r warning=FALSE,message=FALSE}
library(slingshot)
library(plotly)
library(gam)
library(RColorBrewer)
library(biomaRt)
library(monocle)
```

## Read in input data (transcript tpm)

### Read input files. The input file is from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE98664
```{r warning=FALSE}
Hayashi2018 <- read.table("../data/mESC/GSE98664_tpm_sailfish_mergedGTF_RamDA_mESC_differentiation_time_course.txt",
                          sep = "\t", header = T, row.names = 1)
```

### Identify cell types
```{r}
HayashiCellTypes <- data.frame()
for (col in colnames(Hayashi2018)){
HayashiCellTypes[col,"Type"] <- strsplit(col,"_")[[1]][3]
}
tail(HayashiCellTypes)
```

## Prepare annotation

```{r}
# 从Ensembl数据库中加载了小鼠（mmusculus）的基因信息数据集
mouse = useMart("ensembl", dataset = "mmusculus_gene_ensembl")
```

```{r}
# 从之前加载的Ensembl数据库中提取了基因的三个属性：ensembl_gene_id（Ensembl基因ID）、mgi_symbol（小鼠基因名称）和ensembl_transcript_id（Ensembl转录本ID），并将结果存储在annot变量中。
annot <- getBM(c("ensembl_gene_id", "mgi_symbol", "ensembl_transcript_id"), mart=mouse)
```


### Add mgi symbol for the input genes

```{r}
Hayashi2018Copy <- Hayashi2018
transcriptNamesVersion <- do.call(rbind, strsplit(rownames(Hayashi2018), "[.]"))
rownames(Hayashi2018Copy) <- transcriptNamesVersion[, 1]

t_id <- annot$ensembl_transcript_id
dup <- which(duplicated(t_id))
t_id[dup] <- paste("dup_", 1:length(dup)) # ignore duplicated

id_dict = data.frame(id = 1:nrow(annot), row.names = t_id)
mis <- setdiff(rownames(Hayashi2018Copy), rownames(id_dict))
warning(paste(length(mis),
              " transcript_id not found in annotation, accounting for ",
              round(length(mis)/nrow(Hayashi2018Copy)*100, 2),
              "% id in the matrix", sep = ""))
id_dict <- rbind(id_dict, data.frame(id = rep(NA, length(mis)), row.names = mis))

Hayashi2018Copy$mgi_symbol = annot[id_dict[rownames(Hayashi2018Copy), ], "mgi_symbol"]
```

## Prepare input data

### Aggregate transcripts

```{r}
Hayashi2018Final <- aggregate(Hayashi2018[, rownames(HayashiCellTypes)],
                              by=list(geneID=Hayashi2018Copy$mgi_symbol),
                              FUN=sum)

rownames(Hayashi2018Final) <- Hayashi2018Final[, 1]
Hayashi2018Final <- Hayashi2018Final[-1, -1]
rownames(Hayashi2018Final) <- toupper(rownames(Hayashi2018Final))
```

### Find outliers

```{r}
tmp <- colSums(Hayashi2018Final)
hist(tmp)
```

```{r}
to_rmv <- which(abs(tmp-mean(tmp)) > 5*sd(tmp))
Hayashi2018Final <- Hayashi2018Final[, -to_rmv, drop=FALSE]
HayashiCellTypes <- HayashiCellTypes[-to_rmv, , drop=FALSE]
```

## Save / Read data (YOU CAN START FROM HERE)

```{r}
write.csv(Hayashi2018Final, file = "../data/mESC/geneTPM.csv", quote = FALSE)
write.csv(HayashiCellTypes, file = "../data/mESC/celltype.csv", quote = FALSE)
```

```{r}
Hayashi2018Final <- read.csv("../data/mESC/geneTPM.csv",
                             header = T, row.names = 1)
HayashiCellTypes <- read.csv("../data/mESC/celltype.csv",
                             header = T, row.names = 1)
```

## Downstream processing

### Prepare CDS

```{r warning=FALSE, fig.height=10}
# The data is TPM
cd <- new("AnnotatedDataFrame", data = HayashiCellTypes)
gd <- new("AnnotatedDataFrame",
          data = data.frame(rownames(Hayashi2018Final),
                            row.names = rownames(Hayashi2018Final)))

CDS <- newCellDataSet(as(as.matrix(Hayashi2018Final), "sparseMatrix"),
                      phenoData = cd,
                      featureData = gd,
                      lowerDetectionLimit = 0.1,
                      expressionFamily = tobit(Lower = 0.1))

# Next, use it to estimate RNA counts
rpc_matrix <- relative2abs(CDS, method = "num_genes")

# Now, make a new CellDataSet using the RNA counts
CDS <- newCellDataSet(as(as.matrix(rpc_matrix), "sparseMatrix"),
                      phenoData =  cd,
                      featureData = gd,
                      lowerDetectionLimit = 0.5,
                      expressionFamily = negbinomial.size())
```


```{r}
CDS <- estimateSizeFactors(CDS)
CDS <- estimateDispersions(CDS)
```


```{r}
CDS <- detectGenes(CDS, min_expr = 0.1)
print(head(fData(CDS)))
expressed_genes <- row.names(subset(fData(CDS),
                                    num_cells_expressed >= 0.1*dim(cd)[1]))
```

### Log Expr

```{r}
logExpressionFiltered <- log2(as.matrix(exprs(CDS[expressed_genes,])+1))
```

### PCA

```{r}
# Compute PCA
pcaRes <- prcomp(t(logExpressionFiltered), scale. = FALSE)
rd1 <- as.data.frame(cbind(PC1 = pcaRes$x[,1], PC2 = pcaRes$x[,2], PC3 = pcaRes$x[,3]))

plot_ly(as.data.frame(pcaRes$x), x=~PC1, y= ~PC2,  color = HayashiCellTypes$Type, colors = brewer.pal(5,"Set1"))

write.csv(pcaRes$x[, 1:30], "../data/mESC/pca30.csv", quote = F)
```

### Run slingshot

```{r}
slingshotPT <- slingshot(rd1, reducedDim = rd1,
                         clusterLabels = HayashiCellTypes$Type,
                         start.clus = "00h", end.clus = "72h")
ssPT <- slingPseudotime(slingshotPT)
ssPT <- as.data.frame(ssPT)

plot_ly(rd1, x=~PC1, y=~PC2,  color=ssPT$Lineage1)
```

### Find variable genes along the trajectory

```{r}
t <- ssPT$Lineage1


Y <- logExpressionFiltered
var1K <- names(sort(apply(Y,1,var),decreasing = TRUE))
Y <- Y[var1K, ]

# fit a GAM with a loess term for pseudotime
gam.pval <- apply(Y,1,function(z){
  d <- data.frame(z=z, t=t)
  suppressWarnings({
    tmp <- gam(z ~ lo(t), data=d)
  })
  p <- summary(tmp)[4][[1]][1,5]
  p
})
```


```{r}
topgenes <- names(sort(gam.pval, decreasing = FALSE))[1:15]
heatdata <- logExpressionFiltered[topgenes, order(t, na.last = NA)]
heatclus <- as.factor(HayashiCellTypes[order(t, na.last = NA),])

heatmap(as.matrix(heatdata), Colv = NA,
        ColSideColors = brewer.pal(6,"Set1")[heatclus], labCol = FALSE)
```

### Save results

```{r}
exprData <- logExpressionFiltered
colnames(exprData) <- gsub(pattern = "[.]", replacement = "_", colnames(exprData))

ptData <- data.frame(ssPT$Lineage1)
rownames(ptData) <- colnames(exprData)
colnames(ptData) <- "PseudoTime"

geneData <- data.frame(sort(gam.pval, decreasing = FALSE))
colnames(geneData) <- "VGAMpValue"
geneData[,"Variance"] <- apply(logExpressionFiltered[rownames(geneData),],1,var)


write.csv(x = exprData, file = "../data/mESC/OriginalExpressionData.csv", quote = FALSE)
write.csv(x = ptData, file = "../data/mESC/PseudoTime.csv", quote = FALSE)
write.csv(x = geneData, file = "../data/mESC/GeneOrdering.csv", quote = FALSE)
```

### NEXT, YOU MAY RUN `scripts/02_generateExInputs.py`